Detection of slow slip events using wavelet analysis of GNSS recordings

### 2021-10-24 paper/paper_bssa.tex
Start including comments from DS

\begin{itemize}
\item We use a wavelet-based signal processing method to detect slow slip events in GNSS data.
\item There is a good correlation between events detected with GNSS data and events detected with seismic data.
\item The method could be applied in regions and no tremor are detected in conjonction with slow slip events.
\end{itemize}

@@

Slow slip events were discovered in many subduction zones during the last two decades thanks to recordings of the displacement of Earth's surface by GNSS networks. Slow slip can last from a few days to several years and has a relatively short recurrence time (months to years), compared to the recurrence time of regular earthquakes (up to several hundreds of years). In many places, tectonic tremor is also observed in relation to slow slip and can be used as a proxy to study slow slip events of moderate magnitude where surface deformation is hidden in GNSS noise. However, in subduction zones where no clear relationship between tremor and slow slip occurrence is observed, these methods cannot be applied, and we need other methods to be able to better detect and quantify slow slip. In this paper, we use the Maximal Overlap Discrete Wavelet Transform (MODWT) to analyze GNSS time series and seismic recordings of slow slip events in Cascadia. We use detrended GNSS data, apply the MODWT transform and stack the wavelet details over several nearby GNSS stations. As an independent check on the timing of slow slip events, we also compute the cumulative number of tremor in the vicinity of the GNSS stations, detrend this signal, and apply the MODWT transform. In both time series, we can then see simultaneous waveforms whose timing corresponds to the timing of slow slip events. We assume that there is a slow slip event whenever there is a peak in the wavelet signal. We verify that there is a good correlation between slow slip events detected with only GNSS data, and slow slip events detected with only seismic data. The wavelet-based detection method detects all events of magnitude higher than 6 as determined by independent event catalogs (e.g. ~\citet{MIC_2019}).

@@

Slow slip events are a new feature discovered in the last two decades in many subduction zones thanks to recordings of the displacement of Earth's surface by dense Global Navigation Satellite System (GNSS) networks. As with ordinary earthquakes, slow slip events are caused by slip on a fault, such as the plate boundary between a tectonic plate subducting under another tectonic plate. However, they take a much longer time (several days to several years) to happen relative to ordinary earthquakes, and they have a relatively short recurrence time (months to years), compared to the recurrence time of regular earthquakes (up to several hundreds of years), allowing scientists to observe and study many complete event cycles, which is typically not possible to explore with traditional earthquake catalogs ~\citep{BER_2011}. A slow slip event on the plate boundary is inferred to happen when there is a reversal of the direction of motion at GNSS stations, compared to the secular interseismic motion. Slow slip events have been observed in many subduction zones, such as Cascadia, Nankai (southwest Japan), Alaska, Costa Rica, Mexico, and New Zealand ~\citep{BER_2011,AUD_2016}. \\

In many places, tectonic tremor are also observed in relation to slow slip. Tremor is a long (several seconds to many minutes), low amplitude seismic signal, with emergent onsets, and an absence of clear impulsive phases. Tectonic tremor have been explained as a swarm of small, low-frequency earthquakes (LFEs) ~\citep{SHE_2007_nature}, that is small magnitude earthquakes (M $\sim$ 1), for which frequency content (1-10 Hz) is lower than for ordinary earthquakes (up to 20 Hz). In subduction zones such as Nankai and Cascadia, tectonic tremor observations are spatially and temporally correlated with slow slip observations~\citep{OBA_2002,ROG_2003}. Due to this correlation, these paired phenomena have been called Episodic Tremor and Slip (ETS). However, this is not always the case. For instance, in northern New Zealand, tremor are more challenging to detect, and seem to be located downdip of the slow slip on the plate boundary. \\

In Cascadia and Guerrero, Mexico, tremor has been used as a proxy to observe slow slip events that are not directly detectable in the GNSS data. For instance, ~\citet{AGU_2009} studied 23 ETS events in Cascadia with more than 50 hours of tectonic tremor. For all these events, they computed both the GPS-estimated moment release and the cumulative number of hours of tectonic tremor recorded. They observed a linear relationship between moment release and number of hours of tremor for ETS events of moment magnitude 6.3 to 6.8. They also observed many smaller bursts of tremor of duration 1 to 50 hours in between the big ETS events, without any detectable signal in the GPS data. However, based on the relationship between slow slip moment and number of hours of tremor for bigger events, it is possible to infer the existence of smaller slow slip events of magnitude 5-6 occurring simultaneously with the tremor bursts. \\ %This leads to a power-law relationship between seismic moment and number of events with a $b$-value close to one, similar to the distribution of normal earthquakes (\textcolor{red}{reference?}). \\

~\citet{FRA_2016} divided GPS time series observations into two groups: the first group contains days when slow seismicity (tectonic tremor and LFEs) is detected, the second group contains days when the numbers of tremor or LFEs is lower than a threshold. He then stacked separately the two groups of observations and observed a cumulative displacement in the northern direction (for Guerrero) and the eastern direction (for Cascadia) corresponding to the loading period when few tremor or LFEs are observed and the surface deformation corresponds to the secular plate motion. He also observed a cumulative displacement in the southern direction (for Guerrero) and the western direction (for Cascadia) corresponding to the release period when tremor and LFEs are observed. He was thus able to observe a reverse displacement corresponding to smaller slow slip events not directly observable in the GPS data. \\

@@

Wavelets methods such as the Discrete Wavelet Transform (DWT) are mathematical tools for analyzing time series simultaneously in the time and the frequency domain by observing how weighted averages of a time series vary from one averaging period to the next. Wavelet methods have been widely used for geophysical applications ~\citep{KUM_1997}. However, few studies have used wavelet methods to analyze recordings of slow slip, and their scope was limited to the detection of the bigger (magnitude 6-7) short-term (a few weeks) events ~\citep{SZE_2008,OHT_2010,WEI_2012,ALB_2019}. \\

~\citet{SZE_2008} determined the timing and the amplitude of 34 slow slip events throughout the Cascadia subduction zone between 1997 and 2005. They modeled the GPS time series by the sum of a linear trend, annual and biannual sinusoids representing seasonal effects, Heaviside step functions corresponding to earthquakes and hardware upgrades, and a residual signal. They then applied a Gaussian wavelet transform to the residual time series to get the exact timing of the slow slip at each GPS station. The idea is that the wavelet transform allows us to analyze the signal both in the time and the frequency domains. A sharp change in the signal will be localized and seen at all levels of the wavelet decomposition, contrary to what happens with the periodic sinusoids of the Fourier transform. \\

>>>>

\begin{itemize}
\item We use a wavelet-based signal processing method to detect transients in GNSS data, such as slow slip events.
\item There is a good correlation between detections using GNSS data and independent detections using seismic data.
\item The method could be applied in regions where no tremor are detected in conjunction with slow slip events.
\end{itemize}

@@

In many places, tectonic tremor is observed in relation to slow slip and can be used as a proxy to study slow slip events of moderate magnitude where surface deformation is hidden in Global Navigation Satellite System (GNSS) noise. However, in subduction zones where no clear relationship between tremor and slow slip occurrence is observed, these methods cannot be applied, and we need other methods to be able to better detect and quantify slow slip. Wavelets methods such as the Discrete Wavelet Transform (DWT) and the Maximal Overlap Discrete Wavelet Transform (MODWT) are mathematical tools for analyzing time series simultaneously in the time and the frequency domain by observing how weighted differences of a time series vary from one period to the next. In this paper, we use wavelet methods to analyze GNSS time series and seismic recordings of slow slip events in Cascadia. We use detrended GNSS data, apply the MODWT transform and stack the wavelet details over several nearby GNSS stations. As an independent check on the timing of slow slip events, we also compute the cumulative number of tremor in the vicinity of the GNSS stations, detrend this signal, and apply the MODWT transform. In both time series, we can then see simultaneous waveforms whose timing corresponds to the timing of slow slip events. We assume that there is a slow slip event whenever there is a positive peak followed by a negative peak in the wavelet signal. We verify that there is a good correlation between slow slip events detected with only GNSS data, and slow slip events detected with only seismic data for northern Cascadia. The wavelet-based detection method detects well events of magnitude higher than 6 as determined by independent event catalogs (e.g. ~\citep{MIC_2019}).

@@

Slow slip events are a new feature discovered in the last two decades in many subduction zones thanks to recordings of the displacement of Earth's surface by dense Global Navigation Satellite System (GNSS) networks. As with ordinary earthquakes, slow slip events represent slip on a fault, for instance the plate boundary between a tectonic plate subducting under another tectonic plate. However, they take a much longer time (several days to several years) to happen relative to ordinary earthquakes. They have a relatively short recurrence time (months to years) compared to the recurrence time of regular earthquakes (up to several hundreds of years), allowing scientists to observe and study many complete event cycles, which is typically not possible to explore with traditional earthquake catalogs ~\citep{BER_2011}. A slow slip event on the plate boundary is inferred to happen when there is a reversal of the direction of motion at GNSS stations, compared to the secular interseismic motion. Slow slip events have been observed in many places, such as Cascadia, Nankai (southwest Japan), Alaska, Costa Rica, Mexico, and New Zealand ~\citep{BER_2011,AUD_2016}. \\

In many places, tectonic tremor is also observed in relation to slow slip, but it is more abundant in some places. Tremor is a long (several seconds to many minutes), low amplitude seismic signal, with emergent onsets, and an absence of clear impulsive phases. Tectonic tremor have been explained as a swarm of small, low-frequency earthquakes (LFEs) ~\citep{SHE_2007_nature}, which are small magnitude earthquakes (M $\sim$ 1) for which frequency content (1-10 Hz) is lower than for ordinary earthquakes (up to 20 Hz). In subduction zones such as Nankai and Cascadia, tectonic tremor observations are spatially and temporally correlated with slow slip observations ~\citep{OBA_2002,ROG_2003}. Due to this correlation, these paired phenomena have been called Episodic Tremor and Slip (ETS). However, this is not always the case. For instance, in northern New Zealand, tremor are more challenging to detect, and seem to be located downdip of the slow slip on the plate boundary ~\citep{TOD_2016}. In Alaska, the tremor zone only partially overlaps the long-term slow slip zone and there does not appear to be any temporal correlation between tremor and slow slip occurrence ~\citep{WEC_2016}. \\

In Cascadia and Guerrero, Mexico, tremor has been used as a proxy to observe slow slip events that are not directly observed in the GNSS data. For instance, ~\citet{AGU_2009} studied 23 ETS events in Cascadia with more than 50 hours of tectonic tremor. For all these events, they computed both the GPS-estimated moment release and the cumulative number of hours of tectonic tremor recorded. They observed a linear relationship between moment release and number of hours of tremor for ETS events of moment magnitude 6.3 to 6.8. They also observed many smaller bursts of tremor of duration 1 to 50 hours in between the big ETS events, without any detectable signal in the GPS data. However, based on the relationship between slow slip moment and number of hours of tremor for bigger events, it is possible to infer the existence of smaller slow slip events of magnitude 5-6 occurring simultaneously with the tremor bursts. \\ %This leads to a power-law relationship between seismic moment and number of events with a $b$-value close to one, similar to the distribution of normal earthquakes (\textcolor{red}{reference?}). \\

~\citet{FRA_2016} divided GPS time series observations from Guerrero and Cascadia into two groups: the first group contains days with abundant tremor and LFEs, the second group contains days when the number of tremor or LFEs is lower than a threshold. He then stacked separately the two groups of daily observations and observed a cumulative displacement in the northern direction (for Guerrero) and the eastern direction (for Cascadia) corresponding to the loading period when few tremor or LFEs are observed and the surface deformation corresponds to the secular plate motion. He also observed a cumulative displacement in the southern direction (for Guerrero) and the western direction (for Cascadia) corresponding to the release period when tremor and LFEs are observed. He was thus able to observe a reverse displacement corresponding to smaller slow slip events not directly observable in the GPS data for individual events. \\

@@

Wavelets methods such as the Discrete Wavelet Transform (DWT) are mathematical tools for analyzing time series simultaneously in the time and the frequency domain by observing how weighted differences of a time series vary from one period to the next. Wavelet methods have been widely used for geophysical applications (e.g. ~\citep{KUM_1997}). However, few studies have used wavelet methods to analyze recordings of slow slip, and their scope was limited to the detection of the bigger (magnitude 6-7) short-term (a few weeks) events ~\citep{SZE_2008,OHT_2010,WEI_2012,ALB_2019}. \\

~\citet{SZE_2008} determined the timing and the amplitude of 34 slow slip events throughout the Cascadia subduction zone between 1997 and 2005 using wavelets. They modeled the GPS time series by the sum of a linear trend, annual and biannual sinusoids representing seasonal effects, Heaviside step functions corresponding to earthquakes and hardware upgrades, and a residual signal. They then applied a Gaussian wavelet transform to the residual time series to get the exact timing of slow slip at each GPS station. The idea is that the wavelet transform allows us to analyze the signal both in the time and the frequency domains. A sharp change in the signal will be localized and seen at all levels of the wavelet decomposition, contrary to what happens with the periodic sinusoids of the Fourier transform. \\
